---
title: "Austin-Weather Rainfall Prediction"
author: "Sara-Kaleem"
date: "03/05/2021"
output: html_document
---

#Reading the datafile into R
```{r}
library(dplyr)
library(DataExplorer)
Austin <- read.csv("austin_weather.csv")
View(Austin)
plot_str(Austin)
config <- configure_report(
  global_ggtheme = quote(theme_minimal(base_size = 14))
)
create_report(Austin, y= "Events", config = config)
```

#The irrelevant columns needs to be dropped 
```{r}
df <- Austin[-c(1,11,13,21) ]
```

#Checking the structure of data
```{r}
str(df)
```

#replacing special characters or letters with 0
```{r}
df[df=="-"] <- 0
df$PrecipitationSumInches[df$PrecipitationSumInches == 'T'] <- 0
```


#By seeing the structure we see that many columns are stored as chr which needs to be changed to int type
```{r}
df$DewPointHighF <- as.double(df$DewPointHighF)
df$DewPointAvgF <- as.double(df$DewPointAvgF)
df$DewPointLowF <- as.double(df$DewPointLowF)
df$HumidityHighPercent <- as.double(df$HumidityHighPercent)
df$HumidityAvgPercent <- as.double(df$HumidityAvgPercent)
df$HumidityLowPercent <- as.double(df$HumidityLowPercent)
df$SeaLevelPressureAvgInches <- as.double(df$SeaLevelPressureAvgInches)
df$VisibilityHighMiles <- as.double(df$VisibilityHighMiles)
df$VisibilityAvgMiles <- as.double(df$VisibilityAvgMiles)
df$VisibilityLowMiles <- as.double(df$VisibilityLowMiles)
df$WindHighMPH <- as.double(df$WindHighMPH)
df$WindAvgMPH <- as.double(df$WindAvgMPH)
df$WindGustMPH <- as.double(df$WindGustMPH)
df$PrecipitationSumInches <- as.double(df$PrecipitationSumInches)
str(df)
```

#Checking the summary and missing values of the dataset
```{r}
summary(df)
sum(is.na(df))
```

CORRELATION HEATMAP
```{r}
install.packages('reshape2')
library(reshape2)
melted_ds <- melt(df)
head(melted_ds)

library(ggplot2)

ggplot(data = melted_ds, aes(x=var1, y=var2, fill=value)) + 
  geom_tile(color = "white" )+
scale_fill_gradient2(low = "blue", high = "red", mid = "white", 
   midpoint = 0, limit = c(-1,1), space = "Lab", 
   name="Pearson\nCorrelation") +
  theme_minimal()+ 
 theme(axis.text.x = element_text(angle = 90, vjust = 1, 
    size = 12, hjust = 1))+
 coord_fixed()
```

DATA EXPLORATION
```{r}
#install.packages('DataExplorer')
library(DataExplorer)
plot_str(df)
plot_missing(df)
plot_histogram(df)
plot_density(df)
```

Packages for Feature Selection
```{r}
#install.packages('Boruta')
#install.packages('mlbench')
#install.packages('caret')
#install.packages('randomForest')
#install.packages('e1071')
library(e1071)
#install.packages('caTools')
library(caTools)
library(Boruta)
library(mlbench)
library(caret)
library(randomForest)
```

FEATURE SELECTION
```{r}
set.seed(111)
boruta <- Boruta(PrecipitationSumInches ~ ., data = df, doTrace = 2)
print(boruta)
plot(boruta , las=2, cex.axis=0.5)

#TENTATIVE FIX
bor <- TentativeRoughFix(boruta)
print(bor)
attStats(boruta)
getSelectedAttributes(boruta , withTentative = TRUE)
```

DATA PARTITION
```{r}
set.seed(222)
DT <- sample(2, nrow(df), replace = T, prob = c(0.7 , 0.3))
train <- df[DT == 1,]
test <- df[DT == 2,]
```


DECISION TREE - EXPERIMENT 1
```{r}
library(rpart)
set.seed(333)
fit1 <- rpart(PrecipitationSumInches ~ ., 
             method = "anova", data = train)
#Plotting DT
plot(fit1, uniform = TRUE,
          main = "Precipitation Decision Tree")
text(fit1, use.n = TRUE, cex = .9)
  
print(fit1)

```

USING THE RPART.PLOT PACKAGE
```{r}
#install.packages('rpart.plot')
library(rpart.plot)
rpart.plot(fit1)
plotcp(fit1)
printcp(fit1)
```

Calculating the RMSE and MAE
```{r}
library(caret)
p <- predict(fit1, test)
RMSE(p, test$PrecipitationSumInches)
q <- predict (fit1 , train)
RMSE(q, train$PrecipitationSumInches)

```


DECISION TREE WITH DIFFERENT SPLIT RATIO - EXPERIMENT 2
```{r}
set.seed(224)
DT1 <- sample(2, nrow(df), replace = T, prob = c(0.8 , 0.2))
trainDT <- df[DT1 == 1,]
testDT <- df[DT1 == 2,]
set.seed(322)
fit3 <- rpart(PrecipitationSumInches ~ ., 
             method = "anova", data = trainDT)
print(fit3)
rpart.plot(fit3)
plotcp(fit3)
pred_DT <- predict(fit3, testDT)
RMSE(pred_DT, testDT$PrecipitationSumInches)
pred_DT1 <- predict (fit3 , trainDT)
RMSE(pred_DT1, trainDT$PrecipitationSumInches)

```


SUPPORT VECTOR REGRESSION using RBF KERNEL - EXPERIMENT 1
```{r}
library(caTools)
set.seed(123)
split = sample.split(df$PrecipitationSumInches, SplitRatio = 0.7)
training_set = subset(df, split == TRUE)
test_set = subset(df, split == FALSE)


# ~~~~~~~~~~~~~~~~~~~~  Default SVM Model using the RBF kernel ~~~~~~~~~~~~~~~~~~~~~
library(e1071)
svm_rbf <- svm(PrecipitationSumInches~., data = training_set)
summary(svm_rbf)

pred = predict (svm_rbf, test_set)
pred
pred1 = predict (svm_rbf , training_set)
library(caret)
summary(pred)
RMSE(pred, test_set$PrecipitationSumInches) # Root mean squared error
MAE(pred, test_set$PrecipitationSumInches)

RMSE(pred1, training_set$PrecipitationSumInches) # Root mean squared error

```


SUPPORT VECTOR REGRESSION USING FEATURE SELECTION - EXPERIMENT 2
```{r}
set.seed(121)
split = sample.split(df$PrecipitationSumInches, SplitRatio = 0.7)
training_set2 = subset(df, split == TRUE)
test_set2 = subset(df, split == FALSE)


# ~~~~~~~~~~~~~~~~~~~~  Default SVM Model using the RBF kernel ~~~~~~~~~~~~~~~~~~~~~
library(e1071)
svm_rbf1 <- svm(PrecipitationSumInches~ TempHighF +  TempAvgF +  TempLowF +             DewPointHighF + DewPointAvgF + DewPointLowF +             HumidityAvgPercent + HumidityLowPercent +       SeaLevelPressureAvgInches + VisibilityAvgMiles +      VisibilityLowMiles + WindHighMPH + WindAvgMPH +                WindGustMPH , data = training_set2)
summary(svm_rbf1)

pred_svm = predict (svm_rbf1, test_set2)
pred_svm
pred1_svm = predict (svm_rbf1 , training_set2)
library(caret)
summary(pred_svm)
RMSE(pred_svm, test_set2$PrecipitationSumInches) # Root mean squared error

RMSE(pred1_svm, training_set2$PrecipitationSumInches) # Root mean squared error

```


DATA NORMALIZATION - FEATURE SCALING
```{r}
# load libraries
library(caret)
# summarize data
summary(df)
# calculate the pre-process parameters from the dataset
preprocessParams <- preProcess(df, method=c("range"))
# summarize transform parameters
print(preprocessParams)
# transform the dataset using the parameters
transformed <- predict(preprocessParams, df)
# summarize the transformed dataset
summary(transformed)
```


SUPPORT VECTOR REGRESSION USING FEATURE SCALING - EXPERIMENT 3
```{r}
set.seed(321)
split = sample.split(transformed$PrecipitationSumInches, SplitRatio = 0.7)
training_set1 = subset(transformed, split == TRUE)
test_set1 = subset(transformed, split == FALSE)


# ~~~~~~~~~~~~~~~~~~~~  Default SVM Model using the RBF kernel ~~~~~~~~~~~~~~~~~~~~~
library(e1071)
svm_rbf2 <- svm(PrecipitationSumInches~., data = training_set1)
summary(svm_rbf2)

pred3 = predict (svm_rbf2, test_set1)
pred4 = predict (svm_rbf2 , training_set1)
library(caret)
summary(pred3)
RMSE(pred3, test_set1$PrecipitationSumInches) # Root mean squared error
RMSE(pred4, training_set1$PrecipitationSumInches) # Root mean squared error

```


SUPPORT VECTOR REGRESSION (FEATURE SCALING & SELECTION) - EXPERIMENT 4
```{r}
library(e1071)
svm_rbf3 <- svm(PrecipitationSumInches~ TempHighF +  TempAvgF +  TempLowF +             DewPointHighF + DewPointAvgF + DewPointLowF +             HumidityAvgPercent + HumidityLowPercent +       SeaLevelPressureAvgInches + VisibilityAvgMiles +      VisibilityLowMiles + WindHighMPH + WindAvgMPH +                WindGustMPH , data = training_set1)
summary(svm_rbf3)

pred3_svm = predict (svm_rbf3, test_set1)
pred4_svm = predict (svm_rbf3 , training_set1)
library(caret)
summary(pred3_svm)
RMSE(pred3_svm, test_set1$PrecipitationSumInches) # Root mean squared error

RMSE(pred4_svm, training_set1$PrecipitationSumInches) # Root mean squared error

```


RANDOM FOREST - EXPERIMENT 1
```{r}
set.seed(232)
RF <- sample(2, nrow(df), replace = T, prob = c(0.7 , 0.3))
train2 <- df[RF == 1,]
test2 <- df[RF == 2,]

library(randomForest)
set.seed(323)
rf1 <- randomForest(PrecipitationSumInches~ ., data = train2 )
#RMSE and MAE with default RF (ntree=500)
rf.test1<-predict(rf1,newdata = test2)
RMSE(rf.test1,test2$PrecipitationSumInches)

rf.train1 <- predict(rf1,newdata = train2)
RMSE(rf.train1 , train2$PrecipitationSumInches)
#PLOTTING RF
plot(rf1)
varImpPlot(rf1)
importance(rf1)
```

RANDOM FOREST TUNING (mtry) - EXPERIMENT 2
```{r}
library(caret)
str(train2)
tuneRF(train2[ ,-17], train2$PrecipitationSumInches,
      stepFactor=0.5,
      plot = TRUE,
      ntreeTry = 400,
      trace = TRUE,
      improve = 0.05)
set.seed(434)
rf22 <- randomForest(PrecipitationSumInches~.,data = train2,
                   ntreeTry = 500, 
                   mtry=5,
                   importance = TRUE,
                   proximity = TRUE)
print(rf22)
#TUNED RF RMSE 
p1 <- predict(rf22, train2)
RMSE(p1, train2$PrecipitationSumInches)
p2 <- predict(rf22, test2)
RMSE(p2, test2$PrecipitationSumInches)

```

Number of Trees to have lowest error
```{r}
which.min(rf1$mse)
``` 

RANDOM FOREST MODEL WITH 62 TREES - EXPERIMENT 3
```{r}
set.seed(121)
rf.62<-randomForest(PrecipitationSumInches~.,data = train2,ntree=62)
rf.62
#RMSE for Random Forest (ntree = 62)
rf.test<-predict(rf.62,newdata = test2)
RMSE(rf.test,test2$PrecipitationSumInches)
MAE(rf.test,test2$PrecipitationSumInches)
rf.train<-predict(rf.62,newdata = train2)
RMSE(rf.train,train2$PrecipitationSumInches)

```




